---
title: "aflpredictor"
format: html
editor: visual
---

library(fitzRoy)
library(tidyverse)
library(tidymodels)
library(lubridate)
library(xgboost)

install.packages("shiny")
library(shiny)


```{r}
# ---- 1. PULL DATA ----
match_results <- fetch_results_afltables(2015:2024)
player_stats <- fetch_player_stats_afltables(2015:2024)

glimpse(match_results)
glimpse(player_stats)
```

```{r}
# ---- 2. FEATURE ENGINEERING ----

# Filter regular season only
match_data <- match_results %>%
  filter(Round.Type == "Regular") %>%
  arrange(Date) %>%
  mutate(Home.Win = ifelse(Margin > 0, 1, 0))

# Calculate rolling form (last 5 games) for each team
get_rolling_form <- function(team_name, data, n = 5) {
  home_games <- data %>%
    filter(Home.Team == team_name) %>%
    mutate(Team = team_name, Win = ifelse(Margin > 0, 1, 0),
           Points.For = Home.Points, Points.Against = Away.Points)
  
  away_games <- data %>%
    filter(Away.Team == team_name) %>%
    mutate(Team = team_name, Win = ifelse(Margin < 0, 1, 0),
           Points.For = Away.Points, Points.Against = Home.Points)
  
  bind_rows(home_games, away_games) %>%
    arrange(Date) %>%
    mutate(
      Rolling.Wins    = zoo::rollmean(Win, n, fill = NA, align = "right"),
      Rolling.PtsFor  = zoo::rollmean(Points.For, n, fill = NA, align = "right"),
      Rolling.PtsAga  = zoo::rollmean(Points.Against, n, fill = NA, align = "right")
    ) %>%
    select(Date, Team, Rolling.Wins, Rolling.PtsFor, Rolling.PtsAga)
}

# Install zoo if needed
if (!require(zoo)) install.packages("zoo")
library(zoo)

# Get form for all teams
all_teams <- unique(c(match_data$Home.Team, match_data$Away.Team))

team_form <- map_dfr(all_teams, ~get_rolling_form(.x, match_data))

# Join form back to match data
model_data <- match_data %>%
  left_join(team_form, by = c("Date" = "Date", "Home.Team" = "Team")) %>%
  rename(Home.Rolling.Wins   = Rolling.Wins,
         Home.Rolling.PtsFor = Rolling.PtsFor,
         Home.Rolling.PtsAga = Rolling.PtsAga) %>%
  left_join(team_form, by = c("Date" = "Date", "Away.Team" = "Team")) %>%
  rename(Away.Rolling.Wins   = Rolling.Wins,
         Away.Rolling.PtsFor = Rolling.PtsFor,
         Away.Rolling.PtsAga = Rolling.PtsAga) %>%
  drop_na() %>%
  mutate(Home.Win = as.factor(Home.Win))

glimpse(model_data)
```

# ---- 3. BUILD THE MODEL ----

# Select features
model_features <- model_data %>%
  select(
    Home.Win,
    Home.Rolling.Wins,
    Home.Rolling.PtsFor,
    Home.Rolling.PtsAga,
    Away.Rolling.Wins,
    Away.Rolling.PtsFor,
    Away.Rolling.PtsAga
  )

# Train/test split (80/20)
set.seed(42)
split <- initial_split(model_features, prop = 0.8, strata = Home.Win)
train <- training(split)
test  <- testing(split)

# Define model
xgb_model <- boost_tree(
  trees = 500,
  tree_depth = 4,
  learn_rate = 0.01
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

# Recipe
recipe <- recipe(Home.Win ~ ., data = train)

# Workflow
wf <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(xgb_model)

# Train
fit <- wf %>% fit(data = train)

# Evaluate on test set
predictions <- fit %>%
  predict(test) %>%
  bind_cols(predict(fit, test, type = "prob")) %>%
  bind_cols(test)

# Metrics
metrics <- predictions %>%
  metrics(truth = Home.Win, estimate = .pred_class)

print(metrics)

# Confusion matrix
conf_mat(predictions, truth = Home.Win, estimate = .pred_class)
``

```{r}
# ---- 4. PLAYER PERFORMANCE MODEL ----

# Aggregate player stats per game
player_model_data <- player_stats %>%
  filter(!is.na(Disposals), !is.na(Goals)) %>%
  arrange(ID, Date) %>%
  group_by(ID, Player) %>%
  mutate(
    # Rolling averages (last 5 games per player)
    Roll.Disposals = zoo::rollmean(Disposals, 5, fill = NA, align = "right"),
    Roll.Goals     = zoo::rollmean(Goals, 5, fill = NA, align = "right"),
    Roll.Tackles   = zoo::rollmean(Tackles, 5, fill = NA, align = "right"),
    Roll.Marks     = zoo::rollmean(Marks, 5, fill = NA, align = "right"),
    
    # Target — next game disposals
    Next.Disposals = lead(Disposals),
    Next.Goals     = lead(Goals)
  ) %>%
  ungroup() %>%
  drop_na(Roll.Disposals, Next.Disposals)

glimpse(player_model_data)
```


```{r}
# ---- 5. TRAIN PLAYER MODEL ----

player_features <- player_model_data %>%
  select(
    Next.Disposals,
    Roll.Disposals,
    Roll.Goals,
    Roll.Tackles,
    Roll.Marks,
    Age,
    Career.Games,
    Time.on.Ground
  ) %>%
  drop_na()

# Train/test split
set.seed(42)
player_split <- initial_split(player_features, prop = 0.8)
player_train <- training(player_split)
player_test  <- testing(player_split)

# Define regression model
xgb_player <- boost_tree(
  trees = 500,
  tree_depth = 4,
  learn_rate = 0.01
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")

# Recipe
player_recipe <- recipe(Next.Disposals ~ ., data = player_train)

# Workflow
player_wf <- workflow() %>%
  add_recipe(player_recipe) %>%
  add_model(xgb_player)

# Train
player_fit <- player_wf %>% fit(data = player_train)

# Evaluate
player_preds <- player_fit %>%
  predict(player_test) %>%
  bind_cols(player_test)

# Metrics
player_preds %>%
  metrics(truth = Next.Disposals, estimate = .pred)
```


```{r}
# ---- 6. PREDICTION FUNCTIONS ----

# Predict match winner
predict_match <- function(home_team, away_team) {
  
  # Get latest form for each team
  home_form <- team_form %>%
    filter(Team == home_team) %>%
    arrange(desc(Date)) %>%
    slice(1)
  
  away_form <- team_form %>%
    filter(Team == away_team) %>%
    arrange(desc(Date)) %>%
    slice(1)
  
  # Build input
  new_match <- tibble(
    Home.Rolling.Wins   = home_form$Rolling.Wins,
    Home.Rolling.PtsFor = home_form$Rolling.PtsFor,
    Home.Rolling.PtsAga = home_form$Rolling.PtsAga,
    Away.Rolling.Wins   = away_form$Rolling.Wins,
    Away.Rolling.PtsFor = away_form$Rolling.PtsFor,
    Away.Rolling.PtsAga = away_form$Rolling.PtsAga
  )
  
  # Predict
  prob <- predict(fit, new_match, type = "prob")
  
  cat("=================================\n")
  cat(home_team, "vs", away_team, "\n")
  cat("=================================\n")
  cat("Home win probability: ", round(prob$.pred_1 * 100, 1), "%\n")
  cat("Away win probability: ", round(prob$.pred_0 * 100, 1), "%\n")
  cat("Prediction:", ifelse(prob$.pred_1 > 0.5, home_team, away_team), "to win\n")
}

# Predict player disposals
predict_player <- function(player_name) {
  
  player_latest <- player_model_data %>%
    filter(Player == player_name) %>%
    arrange(desc(Date)) %>%
    slice(1) %>%
    select(Roll.Disposals, Roll.Goals, Roll.Tackles, 
           Roll.Marks, Age, Career.Games, Time.on.Ground)
  
  if (nrow(player_latest) == 0) {
    cat("Player not found! Check spelling.\n")
    return(NULL)
  }
  
  pred <- predict(player_fit, player_latest)
  
  cat("=================================\n")
  cat("Player:", player_name, "\n")
  cat("=================================\n")
  cat("Predicted disposals next game:", round(pred$.pred, 1), "\n")
}

# ---- TRY IT OUT ----
predict_match("Richmond", "Carlton")
predict_match("Collingwood", "Melbourne")

predict_player("Patrick Cripps")
predict_player("Clayton Oliver")
```





